---
title:  Multilevel Structural Equation Modeling | Exercises 2
author: A. Stefanelli^[[Alberto Stefanelli](https://albertostefanelli.com/)], B. Castanho Silva^[[Bruno Castanho Silva](https://bcastanho.github.io)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
  self_contained: true
thumbnails: true
lightbox: true
gallery: false
highlight: tango
use_bookdown: true
fig_width: 12
fig_height: 8
# header-includes
---
  <style type="text/css">
  #content {
  max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
  }
#table-of-contents {
width: 300px !important;
}

#postamble {
font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>
  

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
                      fig.show = 'hold', 
                      fig.align = "center", 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")
#BiocManager::install("rhdf5")

```



# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}

# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("readr")                 # data import
# install.packages("dplyr")                 # data manipulation
# install.packages("stringr")               # string manipulation

### MODELING ###
# install.packages("MplusAutomation")       # Talk to Mplus from R

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models


# Load the packages 

### DATA MANIPULATION ###
library("readr")      
library("dplyr")      
library("stringr")
library("fastDummies")

### MODELING ###
library("MplusAutomation")       

### VISUALIZATION ###
library("texreg")
library("tidySEM")

```


# Dataset 

The data set used is the 7th wave of the World Values Survey. This wave is taking place in 2017-2021.

Codebook:

- **Q182**   Justifiable: Homosexuality (1. Never justifiable -- 10. Always justifiable)
- **Q184**   Justifiable: Abortion (1. Never justifiable -- 10. Always justifiable)
- **Q185**   Justifiable: Divorce (1. Never justifiable -- 10. Always justifiable)
- **Q186**   Justifiable: Sex before marriage (1. Never justifiable -- 10. Always justifiable)
- **Q188**   Justifiable: Euthanasia (1. Never justifiable -- 10. Always justifiable)
- **Q275**   Education (0. early childhood/No education -- 8. Doctoral or equivalent, ISCED)
- **Q172P**  How often you pray (1. Never, practically never -- 8. Several times a day)


# Exercise instructions

The input example are in square brackets. You can take a look at the .inp on https://github.com/albertostefanelli/MLSEM_exercises if you need some tips on how to fit the models in Mplus.  

1. Fit a unidimensional two-levels factor model measuring attitudes towards progressive values (Q184, Q185, Q186, Q188) with the same factor structure on both levels. Fix one of the loadings for a latent variable to 1, and allow the factor variance to be estimated. [prog_c_model.inp]
4. Perform a partially saturated model fit evaluation to assess at which level there is the highest lack of fit. You can use the partially saturated model example prog_sat_w_model.inp [prog_sat_w_model.inp, prog_sat_b_model.inp]
5. Test for cross-level invariance by imposing equality constraints on the estimated factor loadings at both level. [prog_inv_model.inp]
6. Fit a model with Random Factor Loadings. Interpret the results and compare to the previous model using different in  Deviance (DIC). [prog_random_model.inp]
3. Alternative specification for configural model: allow all factor loadings to be estimated, and fix factor variances to 1. **(OPTIONAL)** [prog_c_alt_model.inp]
1. Assess and interpret the ICCs of the progressive values items (Q184, Q185, Q186, Q188) by fitting a null model (no substantive covariates of the outcome at either level of the model). **(OPTIONAL)** [prog_null_model.inp]

# Mplus Demo using MplusAutomation

Using the demo version with with MplusAutomation is **possible**. Two options 

1. Rename the Mplus folder and executable
  - On WINDOWS Mplus Demo is usually located in "C:\\Program Files\\Mplus Demo". The folder should berenamed as *Mplus*
  - On WINDOWS Mplus Demo executable is called mpdemo.exe. This should be renamed as *mplus.exe*
  - On MAC Mplus Demo is located in /Applications/MplusDemo/. The folder should b erenamed as *Mplus*
  - On MAC Mplus Demo executable is called mpdemo. This should be renamed as *Mplus*
  - On LINUX : )

2. Point mplusModeler to the right folder using Mplus_command argument. 

NB: Due to a bug in MplusAutomation 0.8 varwarnings should be always specified and set to TRUE such as `varwarnings = TRUE`



```{r, echo=F, message=FALSE, warning=FALSE, results='hide', eval=FALSE}


# read and prepare the data ##
wvs7 <- read_csv("https://github.com/albertostefanelli/MLSEM_exercises/raw/master/wvs7_ex_1.csv")
# transform dummy as factor as required by MplusAutomation  
wvs7_f <- wvs7 %>% 
          mutate_at(vars(Country, female), as.factor)
# remove feamle (need it later)
wvs7_f_m <- wvs7_f %>% select(-female)
# write the Mplus model

equality_m_model <- mplusObject(
  TITLE = "2:1:1 Mediation model with Gender Job Equality (Q31P) as mediator;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "BETWEEN = v2x_gender;", # variable with variance ONLY at between level ß
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  ! within level aka LV 1 
  %WITHIN% 
  Q33P ON Q31P;
  
  ! between level aka LV 2 
  %BETWEEN%
  Q33P ON v2x_gender;
  Q31P ON v2x_gender;

  ",
  rdata = wvs7_f_m
)


```



```{r, echo=T, message=FALSE, warning=FALSE, results='hide', eval=FALSE}

## Run the model ##

equality_m_results <- mplusModeler(object = equality_m_model,         # the model syntax
                                   modelout = "equality_m_model_demo.inp", # output file 
                                   run = 1L,                          # how many models should be run (1)
                                   hashfilename = FALSE,              # add a hash of the raw data
                                   varwarnings = TRUE,                # THIS SHOULD BE SET TO TRUE WITH mplusAutomation 0.8 
                                   Mplus_command =  "/Applications/MplusDemo/mpdemo" 
                       )






```

# Preparing data for mplus using R - The 4 big NO

1. No factors 
2. No long names for variables
3. No character variables 
3. No "NAs" only numbers or .


```{r, echo=T, message=FALSE, warning=FALSE}

wvs7 <- read_csv("https://github.com/albertostefanelli/MLSEM_exercises/raw/master/wvs7_ex_2.csv")

### 1. dealing with exogenous dummy variable 
wvs7$gender_sim_with_a_long_name <- round(runif(nrow(wvs7),0,2),0)
table(wvs7$gender_sim_with_a_long_name)

data_mplus <- fastDummies::dummy_cols(wvs7,
  "gender_sim_with_a_long_name",
  remove_first_dummy = TRUE,
  ignore_na = TRUE)
data_mplus

### 2. dealing with long names 
names(data_mplus)[grep("gender_sim_with_a_long_name",names(data_mplus))][-1] <- c(
  "MALE","NONBIN"
  )

### 3. dealing with character variable 
data_mplus$Country <- as.factor(data_mplus$Country)

### 4. dealing with NAs
data_mplus[60:70,c("Country", "Q188")]

prepareMplusData(
      data_mplus,
      "data1.dat",
      dropCols="gender_sim_with_a_long_name"
  )


```

# Null model and ICCs

```{r, echo=T, message=FALSE, warning=FALSE}
# read and prepare the data ##
wvs7 <- read_csv("https://github.com/albertostefanelli/MLSEM_exercises/raw/master/wvs7_ex_2.csv")
# transform dummy as factor as required by MplusAutomation  
wvs7_f <- wvs7 %>% 
          mutate_at(vars(Country), as.factor)
# remove feamle (need it later)
wvs7_f_m <- wvs7_f %>% select(-Q275,-Q172P)

# write the Mplus model
prog_null_model <- mplusObject(
  TITLE = "Null model Q182 Q184, Q185, Q186, Q188;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  Q182 Q184 Q185 Q186 Q188;

  %BETWEEN%
  Q182 Q184 Q185 Q186 Q188;

  ",
  rdata = wvs7_f_m
)

# continue in using MplusAutomation or directly in Mplus 


```