---
title:  Multilevel Structural Equation Modeling | Exercises 1
author: A. Stefanelli^[[Alberto Stefanelli](https://albertostefanelli.com/)], B. Castanho Silva^[[Bruno Castanho Silva](https://bcastanho.github.io)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
  self_contained: true
thumbnails: true
lightbox: true
gallery: false
highlight: tango
use_bookdown: true
fig_width: 12
fig_height: 8
# header-includes
---
  <style type="text/css">
  #content {
  max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
  }
#table-of-contents {
width: 300px !important;
}

#postamble {
font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>
  

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
                      fig.show = 'hold', 
                      fig.align = "center", 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")
#BiocManager::install("rhdf5")

```


# Dataset 

The data set used is the 7th wave of the World Values Survey. This wave is taking place in 2017-2021.

Codebook:

Between-level:

- **v2x_gender** Women political empowerment index 0 to 1 index [V-Dem, 2019]

Within-level:

- **Q31P** Men make better business executives than women do 1 to 5 
- **Q33P** When jobs are scarce, men should have more right to a job than women
- **female**  Assigned sex at birth (most of the countries)

# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}

# Uncomment this once if you need to install the packages on your system 

## DATA MANIPULATION ##
# install.packages("dplyr")                 # data manipulation

## MODELING ##
# install.packages("MplusAutomation")       # Talk to Mplus from R

## TABLES ##
# install.packages("texreg")               # generates regression tables


# Load the packages 

## DATA MANIPULATION ##
library("readr")      
library("dplyr")      

## MODELING ##
library("MplusAutomation")       

## TABLES ##
library("texreg")

```


# Exercises

1. Fit a 2:1:1 model where gender job equality (Q31P) mediates the relationship between women political empowerment (v2x_gender) and gender job rights (Q33P).
2. Assess the estimated Intraclass Correlations (ICCs) for the Y variables. What is their substantive interpretation ?
3. Calculate the ICCs by hand and compare to the Mplus solution.
4. Manually calculate the number of estimated parameters for the 2:1:1 model mediation model. Compare to the Mplus solution. 
5. Remove the covariance between Q33P and Q31P and see how the df change in the Mplus solution **(OPTIONAL)**
6. Manually calculate the indirect effect of women political empowerment (v2x_gender) via gender job equality (Q31P) $a \times b$ and its standard error ($\sigma_{a \times b}$) using Sobel’s formula:
    - Indirect effect = $a \times b$.
    - $\sigma_{a \times b} = \sqrt{b^2 \times \sigma_{a}^2 + a^2 \times \sigma_{b}^2}$.
7. Ask Mplus to calculate indirect and total effect. Compare the coefficient and SE of the indirect effect to your own solution.
8. Interpret the results. Does gender job equality mediates women political empowerment ? Can you interpret this result causally ?
9. Add sex (female) to the model and assess if the results change **(OPTIONAL)** 

# Solutions 

## 1. Fit mediaiton model in Mplus 

```{r, echo=T, message=FALSE, warning=FALSE}
# read and prepare the data ##
wvs7 <- read_csv("https://github.com/albertostefanelli/MLSEM_exercises/raw/master/wvs7_ex_1.csv")
# transform dummy as factor as required by MplusAutomation  
wvs7_f <- wvs7 %>% 
          mutate_at(vars(Country, female), as.factor)
# remove feamle
wvs7_f_m <- wvs7_f %>% select(-female)
# write the Mplus model
# continue in R (MplusAutomation) or directly in Mplus 

```


# References 