---
title:  Multilevel Structural Equation Modeling | Exercises 2
author: A. Stefanelli^[[Alberto Stefanelli](https://albertostefanelli.com/)], B. Castanho Silva^[[Bruno Castanho Silva](https://bcastanho.github.io)]
# date: "`r Sys.Date()`"
# fontsize: 10pt 
bibliography: lib.bib
link-citations: yes
output:
  #pdf_document:
  rmdformats::readthedown:
  self_contained: true
thumbnails: true
lightbox: true
gallery: false
highlight: tango
use_bookdown: true
fig_width: 12
fig_height: 8
# header-includes
---
  <style type="text/css">
  #content {
  max-width: 1500px !important;
/*    !margin-left: 300px !important;
*/
  }
#table-of-contents {
width: 300px !important;
}

#postamble {
font-size: 10px;
}

pre{
  background-color: #FFFFFF;
    font-size: 12px;
}
pre:not([class]) {
  background-color: #D8D8D8;
    color: black;
}

</style>
  

```{r global_options, include=FALSE}
knitr::opts_chunk$set(tidy=FALSE, 
                      fig.show = 'hold', 
                      fig.align = "center", 
                      warning = FALSE, 
                      message = FALSE, 
                      comment = '')
options(width = 300, scipen = 9999)

# last dev version is needed for rmdformats.
# more info https://github.com/juba/rmdformats/issues/92 
# devtools::install_github("juba/rmdformats")
#BiocManager::install("rhdf5")

```


# Dataset 

The data set used is the 7th wave of the World Values Survey. This wave is taking place in 2017-2021.

Codebook:

- **Q182**   Justifiable: Homosexuality (1. Never justifiable -- 10. Always justifiable)
- **Q184**   Justifiable: Abortion (1. Never justifiable -- 10. Always justifiable)
- **Q185**   Justifiable: Divorce (1. Never justifiable -- 10. Always justifiable)
- **Q186**   Justifiable: Sex before marriage (1. Never justifiable -- 10. Always justifiable)
- **Q188**   Justifiable: Euthanasia (1. Never justifiable -- 10. Always justifiable)
- **Q275**   Education (0. early childhood/No education -- 8. Doctoral or equivalent, ISCED)
- **Q172P**  How often to you pray (1. Never, practically never -- 8. Several times a day)

# Environment preparation 

First, let's load the necessary packages to load, manipulate, visualize and analyse the data. 

```{r, echo=T, message=FALSE, warning=FALSE,cache=F}

# Uncomment this once if you need to install the packages on your system 

### DATA MANIPULATION ###
# install.packages("readr")                 # data import
# install.packages("dplyr")                 # data manipulation
# install.packages("stringr")               # string manipulation

### MODELING ###
# install.packages("MplusAutomation")       # Talk to Mplus from R

### VISUALIZATION ###
# install.packages("tidySEM")               # plotting SEM models


# Load the packages 

### DATA MANIPULATION ###
library("readr")      
library("dplyr")      
library("stringr")

### MODELING ###
library("MplusAutomation")       

### VISUALIZATION ###
library("texreg")
library("tidySEM")

```


# Exercise instructions

1. Assess and interpret the ICCs of the progressive values items (Q184, Q185, Q186, Q188) by fitting a null model (no substantive covariates of the outcome at either level of the model). 
1. Fit a unidimensional two-levels factor model measuring attitudes towards progressive values (Q184, Q185, Q186, Q188) with the same factor structure on both levels. Fix one of the loadings for a latent variable to 1, and allow the factor variance to be estimated.
3. Alternative specification: allow all factor loadings to be estimated, and fix factor variances to 1. **(OPTIONAL)**
4. Perform a partially saturated model fit evaluation to assess at which level there is the highest lack of fit.
5. Test for cross-level invariance by imposing equality constraints on the estimated factor loadings at both level.
6. Fit a model with Random Factor Loadings. Interpret the results and compare to the previous model using different in  Deviance (DIC). 

# Null model and ICCs

```{r, echo=T, message=FALSE, warning=FALSE}
# read and prepare the data ##
wvs7 <- read_csv("https://github.com/albertostefanelli/MLSEM_exercises/raw/master/wvs7_ex_2.csv")
# transform dummy as factor as required by MplusAutomation  
wvs7_f <- wvs7 %>% 
          mutate_at(vars(Country), as.factor)
# remove feamle (need it later)
wvs7_f_m <- wvs7_f %>% select(-Q275,-Q172P)

# write the Mplus model
prog_null_model <- mplusObject(
  TITLE = "Null model Q182 Q184, Q185, Q186, Q188;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  Q182 Q184 Q185 Q186 Q188;

  %BETWEEN%
  Q182 Q184 Q185 Q186 Q188;

  ",
  rdata = wvs7_f_m
)


```

```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_null_results <- mplusModeler(object = prog_null_model,           # the model syntax
                                   modelout = "prog_null_model.inp",  # output file 
                                   run = 1L,                          # how many models should be run (1)
                                   hashfilename = FALSE               # add a hash of the raw data
                       )

prog_null_results$results$data_summary$ICC

```

# Configural model with same factor structure across the two levels

```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_c_model <- mplusObject(
  TITLE = "Configural model;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  prog_w BY Q182 (lw1);
  prog_w BY Q184 (lw2);
  prog_w BY Q185 (lw3);
  prog_w BY Q186 (lw4);
  prog_w BY Q188 (lw5);

  %BETWEEN%
  prog_b BY Q182 (lb1);
  prog_b BY Q184 (lb2);
  prog_b BY Q185 (lb3);
  prog_b BY Q186 (lb4);
  prog_b BY Q188 (lb5);
  ",
  rdata = wvs7_f_m
)


```

```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_c_results <- mplusModeler(object = prog_c_model,              # the model syntax
                                   modelout = "prog_c_model.inp",  # output file 
                                   run = 1L,                       # how many models should be run (1)
                                   hashfilename = FALSE            # add a hash of the raw data

                                                      )

```

```{r, echo=T, message=FALSE, warning=FALSE}

prog_c_results$results$summaries$CFI
prog_c_results$results$summaries$TLI
prog_c_results$results$summaries$AIC
prog_c_results$results$summaries$BIC
prog_c_results$results$summaries$RMSEA_Estimate


```

# Configural model alternative specification (OPTIONAL)

```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_c_alt_model <- mplusObject(
  TITLE = "Configural model alternative specification ;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  prog_w BY Q182* (lw1); ! free loading 
  prog_w BY Q184  (lw2);
  prog_w BY Q185  (lw3);
  prog_w BY Q186  (lw4);
  prog_w BY Q188  (lw5);
  prog_w@1;   ! variance of the factor 
  [prog_w@0]; ! mean of the factor 

  %BETWEEN%
  prog_b BY Q182* (lb1); ! free loading 
  prog_b BY Q184  (lb2);
  prog_b BY Q185  (lb3);
  prog_b BY Q186  (lb4);
  prog_b BY Q188  (lb5);
  prog_b@1;   ! variance of the factor 
  [prog_b@0]; ! mean of the factor 
  ",
  rdata = wvs7_f_m
)


```


```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_c_alt_results <- mplusModeler(object = prog_c_alt_model,              # the model syntax
                                   modelout = "prog_c_alt_model.inp",  # output file 
                                   run = 1L,                       # how many models should be run (1)
                                   hashfilename = FALSE            # add a hash of the raw data

                                                      )

```

```{r, echo=T, message=FALSE, warning=FALSE}

params <- prog_c_alt_results$results$parameters$unstandardized
## loadings ## 
loadings_within <- params %>% filter(paramHeader == "PROG_W.BY") %>% 
                   select(est) %>% 
                   unlist() %>%
                   array()


loadings_between <- params %>% filter(paramHeader == "PROG_B.BY") %>% 
                   select(est) %>% 
                   unlist() %>%
                   array()


cbind(loadings_within, loadings_between)
```

# Partially saturated model fit

## Within level

```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_sat_w_model <- mplusObject(
  TITLE = "Partially saturated model within level ;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  Q182 Q184 Q185 Q186 Q188 WITH 
  Q182 Q184 Q185 Q186 Q188;

  %BETWEEN%
  prog_b BY Q182 (lb1);
  prog_b BY Q184 (lb2);
  prog_b BY Q185 (lb3);
  prog_b BY Q186 (lb4);
  prog_b BY Q188 (lb5);
  ",
  rdata = wvs7_f_m
)


```

```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_sat_w_results <- mplusModeler(object = prog_sat_w_model,             
                                   modelout = "prog_sat_w_model.inp",  
                                   run = 1L,                       
                                   hashfilename = FALSE            

                                                      )

```

## Between level

```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_sat_b_model <- mplusObject(
  TITLE = "Partially saturated model between level ;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  prog_w BY Q182 (lw1);
  prog_w BY Q184 (lw2);
  prog_w BY Q185 (lw3);
  prog_w BY Q186 (lw4);
  prog_w BY Q188 (lw5);

  %BETWEEN%
  Q182 Q184 Q185 Q186 Q188 WITH 
  Q182 Q184 Q185 Q186 Q188;

  ",
  rdata = wvs7_f_m
)


```

```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_sat_b_results <- mplusModeler(object = prog_sat_b_model,             
                                   modelout = "prog_sat_b_model.inp",  
                                   run = 1L,                       
                                   hashfilename = FALSE            

                                                      )



```


```{r, echo=T, message=FALSE, warning=FALSE}

comparison_df <- data.frame("Models"= c("Global", "Saturated Within", "Saturated between"), 
                "LL" = c(prog_c_results$results$summaries$LL, 
                         prog_sat_w_results$results$summaries$LL, 
                         prog_sat_b_results$results$summaries$LL),


                "df" = c(prog_c_results$results$summaries$ChiSqM_DF, 
                         prog_sat_w_results$results$summaries$ChiSqM_DF, 
                         prog_sat_b_results$results$summaries$ChiSqM_DF),
      
                "CFI" = c(prog_c_results$results$summaries$CFI, 
                         prog_sat_w_results$results$summaries$CFI, 
                         prog_sat_b_results$results$summaries$CFI),
                
                "TLI" = c(prog_c_results$results$summaries$TLI, 
                         prog_sat_w_results$results$summaries$TLI, 
                         prog_sat_b_results$results$summaries$TLI),
                
                "AIC" = c(prog_c_results$results$summaries$AIC, 
                         prog_sat_w_results$results$summaries$AIC, 
                         prog_sat_b_results$results$summaries$AIC),
                
                "BIC" = c(prog_c_results$results$summaries$BIC, 
                         prog_sat_w_results$results$summaries$BIC, 
                         prog_sat_b_results$results$summaries$BIC),
                
                "RMSEA"= c(prog_c_results$results$summaries$RMSEA_Estimate, 
                         prog_sat_w_results$results$summaries$RMSEA_Estimate, 
                         prog_sat_b_results$results$summaries$RMSEA_Estimate)
)

comparison_df

```

# Cross-level invariance


```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_inv_model <- mplusObject(
  TITLE = "Cross-level invariance model;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; ! multi-level   
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  prog_w BY Q182 (l1);
  prog_w BY Q184 (l2);
  prog_w BY Q185 (l3);
  prog_w BY Q186 (l4);
  prog_w BY Q188 (l5);

  %BETWEEN%
  prog_b BY Q182 (l1);
  prog_b BY Q184 (l2);
  prog_b BY Q185 (l3);
  prog_b BY Q186 (l4);
  prog_b BY Q188 (l5);
  ",
  rdata = wvs7_f_m
)


```

```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_inv_results <- mplusModeler(object = prog_inv_model,            
                                   modelout = "prog_inv_model.inp", 
                                   run = 1L,                      
                                   hashfilename = FALSE         

                                                      )

```



```{r, echo=T, message=FALSE, warning=FALSE}

comparison_df <- data.frame("Models"= c("Configural", "Cross-Level Invariant"), 
                "LL" = c(prog_c_results$results$summaries$LL, 
                         prog_inv_results$results$summaries$LL),

                "df" = c(prog_c_results$results$summaries$ChiSqM_DF, 
                         prog_inv_results$results$summaries$ChiSqM_DF),
      
                "CFI" = c(prog_c_results$results$summaries$CFI, 
                         prog_inv_results$results$summaries$CFI),
                
                "TLI" = c(prog_c_results$results$summaries$TLI, 
                         prog_inv_results$results$summaries$TLI),
                
                "AIC" = c(prog_c_results$results$summaries$AIC, 
                         prog_inv_results$results$summaries$AIC),
                
                "BIC" = c(prog_c_results$results$summaries$BIC, 
                         prog_inv_results$results$summaries$BIC),
                
                "RMSEA"= c(prog_c_results$results$summaries$RMSEA_Estimate, 
                         prog_inv_results$results$summaries$RMSEA_Estimate)
)

comparison_df

```


# Random loadings model



```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_random_model <- mplusObject(
  TITLE = "Cross-level invariance model;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL RANDOM ; ! multi-level with random effects  
               ESTIMATOR = BAYES; ! works only with the bayesian estimator
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  s1 | prog_w BY Q182;
  s2 | prog_w BY Q184;
  s3 | prog_w BY Q185;
  s4 | prog_w BY Q186;
  s5 | prog_w BY Q188;

  %BETWEEN%
  s1-s5 
  prog_b BY Q182;
  prog_b BY Q184;
  prog_b BY Q185;
  prog_b BY Q186;
  prog_b BY Q188;
  ",
  rdata = wvs7_f_m
)


```

```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_random_results <- mplusModeler(object = prog_random_model,            
                                   modelout = "prog_random_model.inp", 
                                   run = 1L,                      
                                   hashfilename = FALSE         
)

```

```{r, echo=T, message=FALSE, warning=FALSE}
params <- prog_random_results$results$parameters$unstandardized
means <- params %>% filter(str_detect(param, "S")) %>%
                filter(str_detect(paramHeader, "Means"))%>% 
                select(param, est,lower_2.5ci,upper_2.5ci)%>%
                dplyr::rename("Loadings Estimates"  = est)

var <- params %>% filter(str_detect(param, "S")) %>%
                filter(str_detect(paramHeader, "Variances"))%>% 
                select(param, est) %>%
                dplyr::rename("Loadings Variance" = est)

table_random_loadings <- left_join(means,var,by="param") %>%
                         select(param,"Loadings Estimates","Loadings Variance", everything() ) %>%
                         dplyr::rename("Parameter" = param) %>%
                         dplyr::rename("Lower CI (95%)" = lower_2.5ci) %>%
                         dplyr::rename("Upper CI (95%)" = upper_2.5ci)

table_random_loadings
```



```{r, echo=T, message=FALSE, warning=FALSE}

# write the Mplus model
prog_c_bayes_model <- mplusObject(
  TITLE = "Baseline invariance model estimated with bayes;",
  VARIABLE = paste(paste(strwrap(paste("USEVARIABLES = ",paste(names(wvs7_f_m), collapse=" "),
                                       ";"), 30), collapse = "\n"),
  "CLUSTER = Country;",    # grouping variable 
  sep = "\n"
  ),
  ANALYSIS = " TYPE = TWOLEVEL; 
               ESTIMATOR = BAYES;
               PROCESSORS = 2;  ! nuber of phisical processor in your machine", 
  MODEL = "
  %WITHIN% 
  prog_w BY Q182;
  prog_w BY Q184;
  prog_w BY Q185;
  prog_w BY Q186;
  prog_w BY Q188;

  %BETWEEN%
  prog_b BY Q182;
  prog_b BY Q184;
  prog_b BY Q185;
  prog_b BY Q186;
  prog_b BY Q188;
  ",
  rdata = wvs7_f_m
)


```


```{r, echo=T, message=FALSE, warning=FALSE, results='hide'}
## Run the model ##

prog_c_bayes_results <- mplusModeler(object = prog_c_bayes_model,            
                                   modelout = "prog_c_bayes_model.inp", 
                                   run = 1L,                      
                                   hashfilename = FALSE         
)

```


```{r, echo=T, message=FALSE, warning=FALSE}

comparison_df <- data.frame("Models"= c("Baseline", "Random Loadings"), 
                "DIC" = c(prog_c_bayes_results$results$summaries$DIC, 
                         prog_random_results$results$summaries$DIC),

                "pD" = c(prog_c_bayes_results$results$summaries$pD, 
                         prog_random_results$results$summaries$pD)

)

comparison_df$delta <-  NA

comparison_df$delta[2] <-  comparison_df$DIC[2] - comparison_df$DIC[1] 

comparison_df
```
The DIC of the model which allows loadings to vary across countries is lower compared to the baseline, indicating better fit. Therefore, allowing factor loadings to vary across countries produces a better fitting model than constraining them to be the same. 


# References 